Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-09-17 manuscript/paper.tex
JT elect vs elastic... our case elastic and elec , squared alpha beta....

\com{INTRO:
The octahedral rotations and tilting in ABO$_3$ perovskites
are usually explained in terms of 
%a tolerance factor that depends on
the relative sizes 
of the three types of ions involved.
}

@@

\com{
SUM:
While B and tilting tend to bring $xy,x^2-y^2$ down in energy,
A and rotations do the opposite.
}

\com{

Rigid octahedral rotations in a perovskite structure 
deform the crystal field ... from the symmetric octahedral ...

AB ions: cubic and octahedral cages....
O ions: octahedral cage....
Cubic structure: octahedral fields of AB and O are aligned... oriented 
Tilting and rotation of the O octahedra.... change their relative orientation....
and the the combined field is no more octahedral but has a lower symmetry. 
AB lattice: V: -10 to 17 percent of V of O lattice....
So... it is a strong perturbation to the octahedral field... and couples eg and t2g manifolds... and in lifts their degeneracies.... 
}

@@

The Jahn--Teller effect~\cite{Jahn-Teller} describes the 
geometric and electronic instability of a %non-linear 
molecule, cluster or crystal with
a symmetric structure and orbitally degenerate ground state
%(i.e., except Kramers degeneracy)
against structural distortions.
The distortions  
reduce the symmetry of the structure
and lift the electronic degeneracy.
This lowers the energy of the system proportional to
the distortion.
The elastic energy of the structure that increases as the square of the distortion
competes with this linear term and the optimum structure 
is obtained at a finite distortion.
As long as the two competing terms have different \az{[dependence, exponent? diff order polynomial?]} on the distortion,
the ground state can have a finite distortion.
We will use this argument and focus only on the changes
 in the electronic energy throughout this paper.

\az{For our main problem, the rigid octahedral rotations in perovskites,
we will see that 
the energies are quadratic with the tilt and rotation angles. 
However, the unit cell volume also changes quadratically with the angles 
so the elastic energy still goes with a doubled power of this ``distortion'',
i.e, it should be quartic in the angles.
So the two will lead to an optimum structure with a finite tilt and rotation.
}

\sm{
Before we present our results,
let's briefly describe 
the kind of behaviour that would indicate a JT instability 
against the tilting and rotations of the octahedra,
and how it relates to the standard JT problem.
As describe in sec.~\ref{sec:JT},
the optimum structure can have a finite distortion
as long as 
the splitting in the single particle spectrum has a different dependence on the distortion
than the elastic energy associated with the changes in the unit cell volume.

%\subsection{Dependence of electronic spectrum on the ``distortion'' angles}
In case of \az{names? Tetragonal distortion, or quadrupolar distortion, }
elongation or compression of a cube of size $a$ along two of its edges, say, by $a\delta_1,a\delta_2$,
the change in the unit cell volume would be linear in $\delta_{1,2}$
as $\sim a^3\left(\delta_1+\delta_1\right)$.
So the elastic energy %(for an isotropic system) 
changes as $\sim\left(\delta_1+\delta_1\right)^2$.
A linear change ($\propto \delta_{1/2}$) in the electronic spectrum is required to produce
 a finite distortion in the optimum structure (or the ground state of full electron-lattice/nuclear problem).
For example, in the case of tetragonal distortion of an octahedron,
 the total energy has a form
$E=-\rho\delta + K \delta^2$, where $\rho$ and $K$ are constants 
describing the changes in the electronic and elastic energies,
and would be optimum at $\delta=\rho/2K$.
In our case, however,
the unit cell volume depends on the ``distortions'' $\alpha,\beta$ as ${v=4a^3 \cos ^2\alpha \cos ^2\beta}$,
and, at $\alpha,\beta\ll1$,
it changes as $-4a^3\left(\alpha ^2+\beta ^2\right)$.
So, the behaviour we are looking for in the electronic spectrum does not have to be linear 
but quadratic, which would compete with a quartic elastic energy $\left[ \sim\left(\alpha ^2+\beta ^2\right)^2 \right]$
 to produce a finite distortion in the optimum structure.

\rev{
There is one point to clarify here. 
We consider single particle spectrum, i.e.,
single electron energy states
to analyse the ground state of many electron system,
which is the lowest energy ``electronic configuration'' 
(a combination of occupied single particle states 
out of all available single particle states, 
or a slater determinant of that combination to be more precise!).
So a degenerate set of states that are completely occupied
makes a non-degenerate many electron state because 
only a single configuration/combination is possible.
For a given number of electrons in the d-shell of the transition metal ion,
we can determine the (most likely) configuration(s) or ground state(s)
using Hund's rule and see if it is degenerate or otherwise.
 }

@@

>>>>

@@

@@

The Jahn--Teller effect~\cite{Jahn-Teller} describes the 
geometric and electronic instability of a %non-linear 
molecule, cluster or crystal with
a symmetric structure and orbitally degenerate ground state
%(i.e., except Kramers degeneracy)
against structural distortions.
The distortions  
reduce the symmetry of the structure
and lift the electronic degeneracy.
This lowers the energy of the system proportional to
the distortion.
The elastic energy of the structure that increases as the square of the distortion
competes with this linear term and the optimum structure 
is obtained at a finite distortion.

%\subsection{Dependence of electronic and elastic energies on the distortion}

The dependence of the elastic energy on the distortion can be roughly explained
as follows.
Consider an 
elongation or a compression of an octahedron along one of its axes by $a\delta$,
where $a$ is the lattice constant of the cubic structure.
The corresponding change in the unit cell volume would be linear in $\delta$
as $\sim a^3 \delta$
so the elastic energy
would change as $\sim\delta^2$.
Assuming a linear change 
 in the electronic spectrum,
  the total energy has a form
$E=-\rho\delta + K \delta^2$, where $\rho$ and $K$ are constants 
describing the changes in the electronic and elastic energies,
and would be optimum at $\delta=\rho/2K$.
In general, as long as the two competing terms 
are not the same degree monomial in the distortion,
the ground state can have a finite distortion.
We will use this argument and focus only on the changes
 in the electronic energy throughout this paper.

\com{
The dependence of the elastic energy on the distortion can be roughly explained
as follows.
Consider a 
hypothetical 
distortion of an octahedron 
along two of its axes, say, by $a\delta_1,a\delta_2$,
where $a$ is the lattice constant of the cubic structure.
The corresponding change in the unit cell volume would be linear in $\delta_{1,2}$
as $\sim a^3\left(\delta_1+\delta_1\right)$
so the elastic energy %(for an isotropic system) 
changes as $\sim\left(\delta_1+\delta_1\right)^2$.
A linear change ($\propto \delta_{1/2}$) in the electronic spectrum is required to produce
 a finite distortion in the optimum structure (or the ground state of full electron-lattice/nuclear problem).
}

It is worth reminding that, in the Jahn-Teller problem,
we usually consider single particle spectrum, i.e.,
single electron energy states,
to analyse the ground state of a many electron system,
which is the lowest energy ``electronic configuration'' 
(a combination of occupied single particle states 
out of all available single particle states, 
or a slater determinant of that combination to be more precise!).
So a degenerate set of states that are completely occupied
makes a non-degenerate many electron state because 
only a single configuration/combination is possible.
For a given number of electrons in the d-shell of the transition metal ion,
we can determine the (most likely) configuration(s) or ground state(s)
using Hund's rule and see if it is degenerate or otherwise.

@@

Before we present our results,
let's briefly describe 
the kind of behaviour that would indicate a JT instability 
against the tilting and rotations of the octahedra,
and how it relates to the standard JT problem.
As describe in sec.~\ref{sec:JT},
the optimum structure can have a finite distortion
as long as 
the splitting in the single particle spectrum has a different dependence on the distortion
than the elastic energy associated with the changes in the unit cell volume.
In contrast to 
%hypothetical 
elongations or compressions 
$a\delta_1,a\delta_2$ 
of an octahedron along its axes,
where the 
 the unit cell volume changes
as $\sim a^3\left(\delta_1+\delta_1\right)$
and thus the elastic energy approximately goes
as $\sim\left(\delta_1+\delta_1\right)^2$,
the unit cell volume,
in case of octahedral rotation and tilting,
 depends on the ``distortions'' $\alpha,\beta$ 
as ${v=4a^3 \cos ^2\alpha \cos ^2\beta}$,
and, at $\alpha,\beta\ll1$,
it changes as $\sim a^3\left(\alpha ^2+\beta ^2\right)$
leading to a quartic elastic energy that should roughly be
$\sim\left(\alpha ^2+\beta ^2\right)^2$.
So, the behaviour we are looking for in the electronic spectrum 
is not linear
but quadratic, 
which would compete with such a quartic elastic energy
 to produce a finite distortion in the optimum structure.
Thinking of squared angles as the basic variables,
the order of the electronic and elastic energy terms
matches to the standard JT problem.
